---
title: "Lab 2 - Quantitative"
author: "PUT YOUR NAME HERE"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=TRUE, warning=FALSE)
library(mosaic)
library(knitr)
```

# Preparation

1.  Use Save As and save your copy of this as **Lab 2 Report.Rmd** (drop "starter").
2.  Fill your name into the Author field above.
3.  Open the document **STA 215 R Sheet ggformula How-To** from Blackboard and look at the Chapter 3 section. Use that reference to do the analyses requested below.
4.  As you make changes, I encourage you to just keep Knitting the entire document rather than running code chunks here in the editor.

# Part 1. Sale Price of Home Sales in Ames, Iowa

Do Textbook **Problem 3.71** (p. 178). The code chunk below reads the data into a dataframe named **ames**. Each of these problems corresponds to a formula in the **CH3: One Quantitative** section of the 215 R sheet.

```{r ames-input}
ames <- read.csv("AmesHousing.csv")
```

(a) In the code chunk below write a command to obtain a five-number summary for the variable **SalePrice**. (Find the **Basic Numerical Summaries** heading.**)**

```{r ames-a}

```

(b) Find the range and the interquartile range of the sales prices (label each).

**ANSWER:**

(c) In the code chunk below write a command to find the 90th percentile of the variable **SalePrice**. (Find the **Percentile** heading.)

```{r ames-plots}

```

(d) Interpret what the 90th percentile tells you about Ames, IA housing sales from 2006 to 2010.

**ANSWER:**

(e) In the code chunk below write a command to get a boxplot of the variable **SalePrice**. (**Boxplot** heading)

```{r, part e}
# boxplot of the variable SalePrice

```

(f) In the code chunk below write a command to make a histogram of the variable **SalePrice** with first bar starting at 0 and last bar ending at 800,000 with jumps of 20,000 between bars. (**Histogram** heading)

```{r, part f}
# histogram of the variable SalePrice

```

(g) Describe the shape of the distribution in terms of symmetry. If you describe it as skewed, include another word or two indicating how much.

**ANSWER:**

(h) Is the data set unimodal, bimodal, or neither of these?

**ANSWER:**

(i) For any peaks, state the approximate sales price value.

**ANSWER:**

(j) Comment on the presence of outliers in the data set.

**ANSWER:**

(k) Find the mean and standard deviation.

**ANSWER:**

(l) What impact do the outliers have on the mean?

**ANSWER:**

(m) What impact do the outliers have on the standard deviation?

**ANSWER:**

------------------------------------------------------------------------

**Moving on:** As you proceed through the rest of this assignment, refer back to commands you did in Part 1 to accomplish the same task. Then Copy, Paste, and Modify (I call this CPM) the command, updating the dataframe and the variable to match what you are working with.

------------------------------------------------------------------------

# Part 2 - Colleges Nationwide and in Michigan

In Part 2 of this lab, you'll work with data downloaded from the College Scorecard (<https://collegescorecard.ed.gov/>), which is is a tool from the U.S. Department of Education for exploring information about colleges and universities. The dataframe **colleges2** has data on 73 variables for 1,732 institutions that meet these criteria:

-   Is a Title IV school (processes federal financial aid)

-   The "predominant undergraduate degree awarded" is at least a bachelor's degree

The code below loads the R dataframe **colleges2** into the workspace.

```{r read-college}
load("colleges2.rda")
```

## 1. Average Net Price of Colleges & Universities Nationwide

(a) Enter two commands in the code chunk below to request basic numerical summaries of **netprice** and a histogram of **netprice.** Then below the code chunk describe the distribution, addressing the three key characteristics: shape, center, and spread.

```{r, colleges partA}

```

**ANSWER:**

## 2. Number of degree-seeking undergraduates (UGDS)

CPM (copy, paste, and modify) the commands from the previous code chunk to request basic summary statistics and a histogram of the column **UGDS**. Add a third command to also produce a boxplot of this column.

```{r, ugds}

```

(a) Describe the distribution of this variable (shape, center, and spread).

**ANSWER:**

(b) Comment on the presence of outliers in the data set.

**ANSWER:**

## 3. Graduation Rate (GradRate8yr)

The variable **GradRate8yr** is the proportion of entering students that graduated at this school within 8 years of entry. Enter two commands in the code chunk below to request basic numerical summaries and a histogram of this column. Then below the code chunk describe the distribution, addressing the three key characteristics: shape, center, and spread.

```{r, gradrate}

```

**ANSWER:**

### Read some remarks about this kind of variable

There is a key observation to make about this variable. For any individual school the value of this variable (a proportion) was computed from a nominal categorical variable (graduated or not). We say the data used within a school are *student-level* data -- a row corresponds to one student.

However, when analyzing the College Scorecard data, the data are *school-level* data -- a row corresponds to one school. Graduation rate functions as a quantitative characteristic of that school. It is common to analyze a percent or rate variable for which each value was a summary statistic computed from a categorical variable at a lower level of analysis.

------------------------------------------------------------------------

## Colleges & Universities in Michigan

### Some data wrangling -- create dataframes MIschools and GV (code provided)

The code below code creates a dataframe named **MIschools** with only the rows for Michigan schools, keeping only a subset of the variables, and ordering the rows by two variables. Then kable() shows the **MIschools** data in a nice looking table. **MIschools** is the dataframe to be analyzed now. 

```{r MI schools}
MIschools <- colleges2 %>% 
  filter(STABBR=="MI") %>%
  select(INSTNM, control, netprice, ACTCMMID, RET_FT4, GradRate8yr, FIRST_GEN) %>%
  arrange(control, INSTNM) 

kable(MIschools)
```

This code creates a one-row dataframe named **GV**. This is used later to create reference lines to show where GVSU falls on variables relative to other schools.

```{r GV}
## Creates a one-row dataframe for GVSU
GV <- MIschools %>% 
  filter(INSTNM=="Grand Valley State University") 
```

### 2. Net Price of Michigan Schools

The code below creates a comparative boxplot of **netprice** by **control** for the **MIschools** table. Control indicates the type of Institution (1=public 2=private).

A horizontal reference line is added for GVSU's average net price. The use of the chaining operator, %\>%, allows us to overlay additional plot elements. In the style of Mad Libs, complete the sentence below by replacing the placeholders with your words.

```{r MI net price}
gf_boxplot(netprice~control, data=MIschools)  %>% 
  gf_hline(yintercept=~GV$netprice, color="darkblue", linetype="dashed", size=1)
```

**INTERPRETATION:** The average net price for GVSU is **[ADVERB] [ADJECTIVE]** among public schools in Michigan, but would be **[ADVERB] [ADJECTIVE]** among private schools.

## Boxplots of Four Variables for Schools in Michigan

In each of the next sections, you are to do two things:

1.  Copy/Paste/Modify (CPM) the command from above to create a comparative boxplot for the specified variable. You'll need to replace the variable name in two places. You must match the case for variable names.
2.  Complete the interpretive sentence that follows by filling in words where indicated.

### 3. Retention Rate (RET_FT4)

```{r MI-ret-rate}

```

**INTERPRETATION:**

-   Among public schools there are two outliers with very high retention rates, but otherwise there is very little variation in retention rate
-   The distribution of retention rate for private schools has a similar median to that of public schools but there is **[ADVERB] [ADJECTIVE]** variation.
-   Among public schools the retention rate at GVSU is **[ADVERB] [ADJECTIVE]**

### 4. ACT Combined Midpoint (ACTCMMID)

```{r ACTCMMID}

```

**INTERPRETATION:**

-   The ACT Midpoint for GVSU is very close to the **[STATISTIC]** of public schools in Michigan
-   The ACT Midpoint for GVSU is very close to the **[STATISTIC]** of private schools in Michigan

### 5. Graduation Rate (GradRate8yr)

```{r MI-Grad}

```

**INTERPRETATION:**

-   The distribution of Graduation Rate is **[TWO WORDS]** for public and private schools in Michigan,
-   Grand Valley's Graduation Rate is higher than that of approximately **[VALUE]** percent of schools in Michigan

### 6. First-Generation (FIRST_GEN)

Proportion of students that are first-generation (GVSU uses a different criteria in publications)

```{r MI-FirstGen}

```

INTERPRETATION:

-   Relative to other public schools in Michigan, GVSU has a **[ADVERB] [ADJECTIVE]** proportion of students that are first-generation.
-   Compared to public schools, there is **[ADVERB] [ADJECTIVE]** variability in the proportion of students that are first-generation at private schools.

# Preparing to Submit

When everything is complete, create a PDF document by selecting the arrow next to Knit, and then **Knit to PDF**. Then download the report document to your computer and upload that to Blackboard.
