---
title: "Lab 2 - Quantitative"
author: "PUT YOUR NAME HERE"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=TRUE, warning=FALSE)
library(mosaic)
library(knitr)
```

# Preparation

1.  Use Save As and save your copy of this as **Lab 2 Report.Rmd** (drop "starter").
2.  Fill your name into the Author field above.
3.  Open the document **STA 215 R Sheet ggformula How-To** from Blackboard and look at the Chapter 3 section. Use that reference to do the analyses requested below.
4.  As you make changes, I encourage you to just keep Kniting the entire document rather than running code chunks here in the editor.

# Part 1. Sale Price of Home Sales in Ames, Iowa

Do Textbook **Problem 3.71** (p. 178). The code chunk below reads the data into a dataframe named **ames**. Each of these problems corresponds to a formula in the **CH3: One Quantitative** section of the 215 R sheet.

```{r ames-input}
ames <- read.csv("AmesHousing.csv")
```

(a) In the code chunk below write a command to obtain a five-number summary for the variable **SalePrice**. (Find the **Basic Numerical Summaries** heading.**)**

```{r ames-a}

```

(b) Find the range and the interquartile range of the sales prices (label each).

**ANSWER:**

(c) In the code chunk below write a command to find the 90th percentile of the variable **SalePrice**. (Find the **Percentile** heading.)

```{r ames-plots}

```

(d) Interpret what the 90th percentile tells you about Ames, IA housing sales from 2006 to 2010.

**ANSWER:**

(e) In the code chunk below write a command to get a boxplot of the variable **SalePrice**. (**Boxplot** heading)

```{r, part e}
# boxplot of the variable SalePrice

```

(f) In the code chunk below write a command to make a histogram of the variable **SalePrice** with first bar starting at 0 and last bar ending at 800,000 with jumps of 20,000 between bars. (**Histogram** heading)

```{r, part f}
# histogram of the variable SalePrice

```

(g) Describe the shape of the distribution in terms of symmetry. If you describe it as skewed, include another word or two indicating how much.

**ANSWER:**

(h) Is the data set unimodal, bimodal, or neither of these?

**ANSWER:**

(i) For any peaks, state the approximate sales price value.

**ANSWER:**

(j) Comment on the presence of outliers in the data set.

**ANSWER:**

(k) Find the mean and standard deviation.

**ANSWER:**

(l) What impact do the outliers have on the mean?

**ANSWER:**

(m) What impact do the outliers have on the standard deviation?

**ANSWER:**

------------------------------------------------------------------------

**Moving on:** As you proceed through the rest of this assignment, refer back to commands you did in Part 1 to accomplish the same task. Then Copy, Paste, and Modify (I call this CPM) the command, updating the dataframe and the variable to match what you are working with.

------------------------------------------------------------------------

# Part 2 - Colleges Nationwide and in Michigan

In Part 2 of this lab, you'll work with data downloaded from the College Scorecard (<https://collegescorecard.ed.gov/>), which is is a tool from the U.S. Department of Education for exploring information about colleges and universities. The dataframe **colleges2** has data on 73 variables for 1,732 institutions that meet these criteria:

-   Is a Title IV school (processes federal financial aid)
-   The "predominant undergraduate degree awarded" is at least a bachelor's degree

## Descriptions of Selected Columns

| COLUMN NAME | DESCRIPTION                                                                                                                                                              |
|------------------|------------------------------------------------------|
| control     | Type of Institution, 1=public 2=private                                                                                                                                  |
| netprice    | Average net price for Title IV institutions                                                                                                                              |
| ACTCMMID    | Midpoint between the 25th and 75th percentile of ACT cumulative scores.                                                                                                  |
| RET_FT4     | Proportion of full-time, first-time, degree/certificate-seeking undergraduates who were enrolled at the institution in the fall 1 year after starting at the institution |
| GradRate8yr | Proportion of entering students that graduated at this school within 8 years of entry                                                                                    |
| FIRST_GEN   | Proportion of students that are first-generation (GVSU uses a different criteria)                                                                                        |

The code below loads the R dataframe **colleges2** into the workspace. Then **colleges2** is "chained" into a dplyr function to select only a subset of variables to keep, and the resulting narrower dataframe replaces the original. **The dataframe to work with in this section is colleges2.**

```{r read-college}
load("colleges2.rda")
colleges2 <- colleges2 %>%
    select(INSTNM, STABBR, control, netprice, ACTCMMID, RET_FT4, GradRate8yr, FIRST_GEN)
```

## 1. Average Net Price of Colleges & Universities Nationwide

(a) Request basic numerical summaries of **netprice** and a histogram of **netprice.**

```{r, colleges partA}

```

(b) Describe the distribution, addressing the three key characteristics: shape, center, and spread.

**ANSWER:**

(c) Below write code to get side-by-side boxplots with **netprice** for the response variable, and **control** for the explanatory variable. (See 215 R sheet - CH 3 - **Boxplot By-Group**.)

```{r, colleges partC}

```

(d) Write a sentence or two comparing the distribution of **netprice** for private vs public schools.

**ANSWER:**

------------------------------------------------------------------------

## Colleges & Universities in Michigan

### Some data wrangling -- create dataframes MIschools and GV (code provided)

The first chained command creates the new dataframe **MIschools**, assigning to it the result of starting from the **colleges2** table and manipulating it in in three ways:

-   filter() returns a subset of rows that meet the condition. Then ...
-   select() returns a subset of columns -- the negation operator drops a column (it would have "MI" for every row here). Then ...
-   arrange() sorts the rows by **control** and then by **INSTNM**.

The second chained command defines a new object named **GV**, containing the one row in **MIschools** that has that value of **INSTNM**. So GV is a dataframe with one row. Then kable() prints the **MIschools** data in a nice looking table.

```{r MI schools}
MIschools <- colleges2 %>% 
  filter(STABBR=="MI") %>%
  select(-STABBR) %>%
  arrange(control, INSTNM) 

## Creates a one-row dataframe for GVSU
GV <- MIschools %>% 
  filter(INSTNM=="Grand Valley State University") 

kable(MIschools)
```

### 2. Net Price of Michigan Schools

The code below creates a comparative boxplot of **netprice** by **control** for the **MIschools** table. A horizontal reference line is added for GVSU's average net price. The use of the chaining operator, %\>%, allows us to overlay additional plot elements. In the style of Mad Libs, complete the sentence below by replacing the placeholders with your words.

```{r MI net price}
gf_boxplot(netprice~control, data=MIschools) %>%
  gf_hline(yintercept=~GV$netprice, color="darkblue", linetype="dashed", size=1)
```

**INTERPRETATION:** The average net price for GVSU is **[ADVERB] [ADJECTIVE]** among public schools in Michigan, but would be **[ADVERB] [ADJECTIVE]** among private schools.

## Boxplots of Four Variables for Schools in Michigan

In each of the next sections, you are to do two things:

1.  Copy/Paste/Modify (CPM) the command from above to create a comparative boxplot for the specified variable. You'll need to change the variable name in two places. You must match the case for variable names.
2.  Complete the interpretive sentence that follows by filling in words where indicated.

### 3. Retention Rate (RET_FT4)

```{r MI-ret-rate}

```

**INTERPRETATION:**

-   Among public schools there are two outliers with very high retention rates, but otherwise there is very little variation in retention rate
-   The distribution of retention rate for private schools has a similar median to that of public schools but there is **[ADVERB] [ADJECTIVE]** variation.
-   Among public schools the retention rate at GVSU is **[ADVERB] [ADJECTIVE]**

### 4. ACT Combined Midpoint (ACTCMMID)

```{r ACTCMMID}

```

**INTERPRETATION:**

-   The ACT Midpoint for GVSU is very close to the **[STATISTIC]** of public schools in Michigan
-   The ACT Midpoint for GVSU is very close to the **[STATISTIC]** of private schools in Michigan

### 5. Graduation Rate (GradRate8yr)

```{r MI-Grad}

```

**INTERPRETATION:**

-   The distribution of Graduation Rate is **[TWO WORDS]** for public and private schools in Michigan,
-   Grand Valley's Graduation Rate is higher than that of approximately **[VALUE]** percent of schools in Michigan

### 6. First-Generation (FIRST_GEN)

Proportion of students that are first-generation (GVSU uses a different criteria in publications)

```{r MI-FirstGen}

```

INTERPRETATION:

-   Relative to other public schools in Michigan, GVSU has a **[ADVERB] [ADJECTIVE]** proportion of students that are first-generation.
-   Compared to public schools, there is **[ADVERB] [ADJECTIVE]** variability in the proportion of students that are first-generation at private schools.

# Preparing to Submit

When everything is complete, create a PDF document by selecting the arrow next to Knit, and then **Knit to PDF**. Then download the report document to your computer and upload that to Blackboard.
